---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# qryflow

<!-- badges: start -->
[![R-CMD-check](https://github.com/christian-million/qryflow/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/christian-million/qryflow/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/christian-million/qryflow/graph/badge.svg)](https://app.codecov.io/gh/christian-million/qryflow)
[![Downloads](https://cranlogs.r-pkg.org/badges/grand-total/qryflow)](https://cran.r-project.org/package=qryflow)
[![CRAN status](https://www.r-pkg.org/badges/version/qryflow)](https://cran.r-project.org/package=qryflow)
<!-- badges: end -->

## Overview

Execute multi-step 'SQL' statements using specially formatted comments that define and control execution.

`qryflow` lets you define **multi-step SQL workflows** using comment-based tags in your SQL code. These tags tell R how to execute each SQL chunk and what to name the results. This allows you to:

- Keep multiple SQL statements in the same file.

- Control how each SQL "chunk" is executed.

- Return results as named R objects.

- Extend behavior using custom tags, parsers, and handlers.

## Install

You can install the released version of `qryflow` from [CRAN](https://cran.r-project.org/) with:

```r
install.packages("qryflow")
```

And the development version from GitHub with:

```r
# install.packages("devtools")
devtools::install_github("christian-million/qryflow")
```

## Example

The code below demonstrates the primary use case for `qryflow`.

Basic Usage:

```{r example}
library(qryflow)

# Connection to In-Memory DB with table populated from mtcars
con <- example_db_connect(mtcars)

sql <- "
-- @exec: drop_cyl_6
DROP TABLE IF EXISTS cyl_6;

-- @exec: prep_cyl_6
CREATE TABLE cyl_6 AS
SELECT *
FROM mtcars
WHERE cyl = 6;

-- @query: df_mtcars
SELECT *
FROM mtcars;

-- @query: df_cyl_6
SELECT *
FROM cyl_6;
"

# Pass tagged SQL to `qryflow`
results <- qryflow(con, sql)

# Access the results from the chunk named `df_cyl_6`
head(results$df_cyl_6)
```

The path to a file containing SQL can also be passed:

```{r}
filepath <- example_sql_path('mtcars.sql')

# Pass tagged SQL to `qryflow`
results <- qryflow(con, filepath)

# Access the results from the chunk named `df_cyl_6`
results$df_cyl_6 |>
  head()
```

```{r, echo=FALSE, include=FALSE}
DBI::dbDisconnect(con)
```

## Additional Learning

Consider the following vignettes for a more in depth understanding:

- Getting Started: Outlines available features, how to use `qryflow`, and provides an operational understanding of how it works (`vignette("getting-started", package = "qryflow")`).

- Advanced Usage: A look under the hood at the objects and classes that power `qryflow` so that you can get more out of the package (`vignette("advanced-qryflow", package = "qryflow")`).

- Extend `qryflow`: A guide to understanding how to implement custom tags, or override the built-in tags, using custom chunk handlers (`vignette("extend-qryflow", package = "qryflow")`).

## Similar Packages

The functionality made available by `qryflow` exists in other packages. However, the scope and implementation of `qryflow` makes it distinct enough to justify a unique package.

I recommend reviewing these other packages to see which works best for your needs. If you feel this list is incomplete, please submit an issue:

- [`sqlhelper`](https://CRAN.R-project.org/package=sqlhelper) provides comprehensive tools for executing parameterized SQL scripts, managing database connections and configurations, supporting spatial data types, and statement-level control within SQL files.

- [`SQLove`](https://cran.r-project.org/package=SQLove) is 'a lightweight R package for handling complex SQL scripts including temp tables, multiple queries, etc.'
